---
title: "alz"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(alz)
```

## Overview

The `{alz}` package provides structured metadata and helper functions for
working with Alzheimer's disease biomarkers. It formalizes key concepts
related to how biomarkers are **measured**, **interpreted**, and
**mapped to pathology**, using relational data tables.

This vignette introduces the four foundational entities in the package:

- `analytes`: the molecular entities being quantified  
- `assays`: the measurement implementations used to quantify analytes  
- `cutpoints`: the thresholds used to interpret assay results  
- `biomarkers`: biologically or clinically meaningful indicators derived from one or more analytes

Each of these concepts plays a distinct role in the biomarker data model.

## Analytes

An *analyte* is the molecular species being quantified. Examples include:

- Aβ42 (amyloid beta 1–42)
- p-tau181 (tau phosphorylated at threonine 181)
- GFAP (glial fibrillary acidic protein)

Analytes are uniquely identified using external ontology references such as
*ChEBI* or *UniProt*, and are independent of how they are measured or
interpreted.

```r
analytes()
```

## Assays

An assay represents a specific measurement implementation used to quantify an
analyte. It combines:

- a technology (e.g. ELISA, Simoa, PET),
- a platform (e.g. INNOTEST, HD-X),
- a vendor (e.g. Fujirebio, Quanterix),
- a matrix (e.g. CSF, plasma),

and quantitation properties (e.g. LLOQ, ULOQ, unit).

Each row in the `assays()` table reflects a specific commercial or
research-grade measurement system.

## Cutpoints

A cut-point defines how a numeric assay result is discretised into categorical
levels — such as negative or positive. It includes:

- the biomarker being classified
- the assay used (optional if general)
- the threshold(s) and associated labels
- intended use (e.g. screening vs diagnosis)

Cut-points enable classification of raw measurements based on published or
validated rules.

```{r}
cutpoints()
```

## Biomarkers

A biomarker is a biologically or clinically meaningful measurement derived
from one or more analytes. Biomarkers may be:

- original: direct assay of an analyte (e.g. p-tau181 in CSF)
- derived: computed from analytes (e.g. Aβ42/Aβ40 ratio, p-tau217 / non-p-tau217)
- imaging-based: e.g. amyloid PET, FDG PET

Each biomarker is mapped to:

- a category: core, non-specific, co-pathology
- a subcategory: A, T1, T2, N, I, V, S
- a modality: fluid, imaging, or hybrid

```{r}
biomarkers()
```

## Relationships

The entities are connected as follows:

- Assays measure analytes
- Biomarkers are defined from one or more analytes
- Cut-points interpret biomarkers, optionally specific to an assay, and with
an intended use in mind or specific validation circumstances.

This relational structure allows users to navigate from molecular data to
clinical interpretation with full transparency.

```{r}
# Example: get all assays for Aβ42
assays() |> dplyr::filter(analyte_id == "CHEBI:64647")
```

